### 2025-09-25 paper/sec/evaluation.tex
Unpack SciGen.
answer to RQ1

To evaluate the system we used a sample of the \SciGen dataset~\citep{moosavi21}. We aggregated the generated
expression in the linguistic categories from Table~\ref{tab:natural-language-forms}.
Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and
without target-value sharing. The results show that providing the target value (text fragment or numerical
value) is crucial for achieving a high success rate. Overall the \InterpretationAgent generates a correct
expression in approximately in 74.9\% of the cases (S.D. 3.0\%).

The system achieved performance above 68\% in comparison, 77.3\% in data retrieval, and 97\% in min/max search tasks.
In contrast, accuracy decreased significantly in expressions requiring differences (20\%) and in ranking tasks (0\%).

Figure~\ref{fig:success_rate_by_complexity} reports the success rate as a function of the number of categories assigned to each expression.
Fluid instructions combining multiple operations were associated with multiple categories.
Results show that the success rate was 62\% for simple expressions (one category), increased to 91\% for two categories,
while no correct expression was generated for expression with 3 categories.

In addition to evaluating performance when the target expression is provided, we also assessed the model’s ability to produce a
correct expression when no reference string was given.
On the \SciGen dataset, the model achieved a 74.9\% success rate when provided with the target string.
Without the target string, performance decreased moderately to 57.1\%, indicating that the model retains substantial
ability to reconstruct valid solutions even without explicit guidance

>>>>

To evaluate RQ1, we used a sample of the \SciGen dataset~\citep{moosavi21}, an open source dataset consisting
of tables from scientific articles and their corresponding descriptions. We aggregated the generated
expression in the linguistic categories from Table~\ref{tab:natural-language-forms}.
Figure~\ref{fig:success_rate_by_category} illustrates the success rate for each category, both with and
without target-value sharing.

\emph{The results show that the system is robust when provided with sufficient guidance but degrades when underspecified.} With the target-value sharing, 
the InterpretationAgent produced correct Fluid expressions in 74.9\%  (S.D. 3.0\%) of cases,
but performance dropped to 57.1\% when the target was withheld. This highlights the system's reliance on explicit cues when resolving ambiguous fragments.

\emph{Performance also varied across linguistic categories.} Success rates exceeded 68\% for comparison, 77.3\% for data retrieval, and 97\% for min/max search tasks.
In contrast, accuracy decreased significantly for expressions requiring differences (20\%) and for ranking tasks (0\%).

The trend for compositional complexity is more nuanced as shown in Figure~\ref{fig:success_rate_by_complexity}, which reports the success rate as a function of the number of categories assigned to each expression: success rates are 62\% for single-category expressions, increase to 91\% when two categories are combined, but collapse to 0\% when three categories are involved. This suggests that \emph{moderate composition can actually aid performance, perhaps by giving the model clearer structural cues, but that complexity beyond a certain threshold overwhelms the synthesis process}.

%The results show that providing the target value (text fragment or numerical
%value) is crucial for achieving a high success rate. Overall the \InterpretationAgent generates a correct
%expression in approximately in 74.9\% of the cases (S.D. 3.0\%).

%Figure~\ref{fig:success_rate_by_complexity} reports the success rate as a function of the number of categories assigned to each expression.
%Fluid instructions combining multiple operations were associated with multiple categories.
%Results show that the success rate was 62\% for simple expressions (one category), increased to 91\% for two categories,
%while no correct expression was generated for expression with 3 categories.

%In addition to evaluating performance when the target expression is provided, we also assessed the model’s ability to produce a
%correct expression when no reference string was given.
%On the \SciGen dataset, the model achieved a 74.9\% success rate when provided with the target string.
%Without the target string, performance decreased moderately to 57.1\%, indicating that the model retains substantial
%ability to reconstruct valid solutions even without explicit guidance
